---
title: "Gaussian graphical models: an extended tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian graphical models: an extended tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 5)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
library(psychnet)
net <- psychnet(data = SRL_GPT, method = "glasso")
```

```{r}
library(package = psychnet)
```

## What a psychological network represents

A psychological network is a statistical representation of relations among
measured variables. In a Gaussian graphical model (GGM), the nodes are variables:
symptoms, questionnaire items, scale scores, behaviours, or other measured
constructs. The edges are partial correlations. An edge therefore does not mean
that two variables are merely correlated in the ordinary bivariate sense. It means
that they remain associated after conditioning on all other variables included in
the network.

This distinction is central to the interpretation of psychological networks
(Epskamp & Fried, 2018; Borsboom et al., 2021; Saqr, Beck & López-Pernas, 2024).
Suppose that cognitive strategy use and self-regulation are connected in a
network that also contains intrinsic value, self-efficacy, and test anxiety. The
edge asks a ceteris paribus question: are cognitive strategy use and
self-regulation still associated among respondents who are comparable on the
other constructs in the model? If so, the edge represents a conditional
dependence. If not, the pair is conditionally independent in the fitted graph.

The absence of an edge is consequently informative. It says that, given the other
variables in the network, the data do not support a direct conditional association
between the two nodes under the fitted model. This does not prove that the
constructs are unrelated in every substantive sense. It means that any observed
zero-order relation between them can be accounted for through other variables in
the network, sampling variation, or both. Interpreting both present and absent
edges is what makes a GGM more than a visual display of ordinary correlations.

## Why and how we regularize

An unregularized partial-correlation network is usually dense. When the sample
correlation matrix is inverted directly, almost every pair of variables receives a
non-zero partial correlation. Some of those partial correlations may reflect
stable conditional associations, but many small estimates will arise because
sample correlations fluctuate. The problem becomes more serious as the number of
variables grows relative to the number of observations, because each edge is
estimated while adjusting for many other variables.

Regularization addresses this by shrinking weak edges. The graphical lasso
(Friedman, Hastie & Tibshirani, 2008) estimates a sparse precision matrix by
penalizing the absolute size of off-diagonal entries. In the partial-correlation
network, this has the practical consequence that some weak conditional
associations are set exactly to zero. The result is a graph that is easier to
read and less dominated by small sample-specific estimates.

Regularization is not a guarantee of truth. It changes the error tradeoff. A
sparser graph typically reduces false-positive edges, but it may also remove weak
associations that are real in the population. The substantive question is
therefore not simply whether an edge appears, but how strong and stable the edge
is, how plausible the model assumptions are, and how the result changes under
reasonable analytic choices.

The extended Bayesian information criterion (EBIC; Foygel & Drton, 2010) is
commonly used to choose the amount of regularization. Its hyperparameter, usually
called gamma, controls an additional penalty for model complexity. A larger gamma
favours sparser networks; a smaller gamma is more permissive. Choosing gamma is
therefore a modelling decision about conservatism. In psychological-network
applications, gamma equal to 0.5 is often used as a default because it gives a
relatively cautious graph, but researchers should still treat weak and borderline
edges as uncertain until accuracy and stability have been examined.

## A worked analysis with SRL_GPT

The `SRL_GPT` data set contains 300 observations of five Motivated Strategies for
Learning Questionnaire construct scores: cognitive strategy use (`CSU`),
intrinsic value (`IV`), self-efficacy (`SE`), self-regulation (`SR`), and test
anxiety (`TA`). Each variable is a mean score on a 1-7 response scale.

```{r}
head(x = SRL_GPT)
```

The network is estimated with `psychnet()`. Here `method = "glasso"` requests the
EBIC-regularized graphical lasso. The method name `"EBICglasso"` is accepted as
an alias, but this vignette uses the shorter spelling.

```{r}
psychnet(data = SRL_GPT, method = "glasso")
```

The edge list is the most useful first output because it states the fitted
partial correlations directly.

```{r}
as.data.frame(x = net)
```

Each row is a conditional association between two constructs. Positive weights
mean that respondents who are higher on one construct tend also to be higher on
the other, after conditioning on the remaining constructs. Negative weights mean
that respondents who are higher on one construct tend to be lower on the other,
again conditionally on the rest of the network.

In this example, the learning-strategy and motivation constructs form a positive
cluster. `CSU` is positively connected to `IV`, `SE`, and `SR`; `IV`, `SE`, and
`SR` are also positively linked. These edges should be read as conditional
relations among construct scores, not as evidence that one construct causes
another. The strongest negative association is between `SR` and `TA`, suggesting
that test anxiety is lower among respondents with higher self-regulation when
cognitive strategy use, intrinsic value, and self-efficacy are held constant.
Several positive edges involving `TA` are small; their substantive interpretation
should be correspondingly cautious.

Centrality summaries describe how a node is positioned in the fitted network.

```{r}
net_centralities(x = net)
```

Node strength is the sum of the absolute edge weights incident on a node. It is
usually the most interpretable and comparatively stable centrality index in
psychological networks. Expected influence is the signed analogue: positive and
negative edges can offset one another. In this data set, `SR` has high strength
because it is connected to several learning constructs and also has a sizeable
negative edge with `TA`. Its expected influence is lower than its strength because
the negative edge offsets its positive connections.

Other centrality indices, especially betweenness and closeness, are often
unstable in psychological networks and should not be overinterpreted without
dedicated stability evidence (Bringmann et al., 2019). For this reason, this
vignette emphasizes strength and expected influence rather than treating every
available graph-theoretic index as equally meaningful.

Predictability provides an absolute complement to centrality. Instead of asking
how connected a node is, it asks how much of the variance in each node can be
accounted for by its neighbours in the network (Haslbeck & Waldorp, 2018).

```{r}
net_predict(x = net)
```

Predictability is useful because two nodes can have similar centrality but differ
in how well the rest of the network explains them. Here the learning-related
constructs are highly predictable from their neighbours, whereas `TA` is much
less predictable. That pattern fits the edge list: test anxiety has one
substantial negative edge with self-regulation and several smaller positive
edges.

The same fitted object can be plotted with `cograph`. The chunk is guarded so the
vignette can still render on systems where the optional plotting package is not
available.

```{r ggm-plot, eval = has_cograph}
cograph::splot(x = net, psych_styling = TRUE)
```

The graph is a visual summary of the edge list. It is useful for orientation, but
the numerical edge table remains the primary object for interpretation. Layout
algorithms place nodes to make the graph readable; distances on the page should
not be interpreted as direct measurements unless the plotting method explicitly
defines them that way.

## Assessing accuracy and replicability

Point estimates are not enough. A reported edge weight is one estimate from one
sample, and centrality indices are functions of many estimated edges. Researchers
should therefore assess how much those estimates vary under resampling.

For edge weights, nonparametric bootstrapping can be used to obtain uncertainty
intervals and to examine whether differences between edges are large relative to
sampling variability. For centrality, case-dropping procedures ask whether the
centrality ordering remains similar when portions of the sample are removed. This
is the logic behind the bootstrap and stability procedures recommended by
Epskamp, Borsboom, and Fried (2018).

In `psychnet`, `net_boot()` and `net_stability()` provide these analyses. They are
not run here because a thorough bootstrap is computationally longer than a
compact vignette should require, but they should be part of a complete empirical
analysis when edge accuracy or centrality claims are substantively important.

## Non-normal and ordinal data

GGMs are Gaussian models, yet psychological data are often skewed, bounded, and
measured on ordinal response scales. Scale means, such as the five `SRL_GPT`
construct scores, are often treated as approximately continuous, but this is a
modelling approximation rather than a property guaranteed by the questionnaire.

For ordinal item-level data, `cor_method = "auto"` can be used so that the
correlation input is based on polychoric or polyserial associations when
appropriate. This choice is most relevant when variables are individual Likert
items or mixed ordinal-continuous measures rather than scale means.

For continuous variables with pronounced monotone departures from normality,
`method = "huge"` estimates a nonparanormal graphical model. The nonparanormal
approach assumes that variables become jointly Gaussian after unknown monotone
transformations of their marginal distributions (Liu, Lafferty & Wasserman,
2009). It is a way to retain a GGM-style conditional-dependence interpretation
while relaxing strict multivariate normality.

## Native estimation

By default (`native = TRUE`) `psychnet` estimates the graphical lasso with its own
solver, written entirely in base R. This is a *clean-room* implementation: the
algorithm is reconstructed from the original description (Friedman, Hastie &
Tibshirani, 2008) rather than wrapping an existing library, and it reproduces the
estimate from first principles. Its advantages are:

- it depends only on base R, so it installs and runs anywhere with no compilation
  or external library;
- the estimator is plain, readable R code that can be inspected and audited end
  to end, not a compiled black box;
- each fit is graded against the convex objective it optimizes, so its
  correctness is established internally rather than by agreement with another
  program;
- the results are reproducible across platforms.

## The Fortran glasso estimator

The native solver is not the only option. Setting `native = FALSE` delegates
each penalized solve to the compiled `glasso` Fortran routine — the same code
underlying `qgraph::EBICglasso()` — which is faster on large problems and returns
numerically equivalent estimates:

```{r}
psychnet(data = SRL_GPT, method = "glasso", native = FALSE)
```

The native (`native = TRUE`) and Fortran (`native = FALSE`) solvers estimate the
same model; they differ only in the implementation that performs the computation,
so a study can use the convenient native solver and confirm it against the
established external one.

## Mathematical foundations

This final section gives the statistical definitions behind the tutorial. It is
included for readers who want the mathematical form of the model and estimator,
and can be skipped without losing the worked example above.

### Precision matrix and the pairwise Markov property

Let $\mathbf{X} = (X_1, \ldots, X_p)$ follow a multivariate normal distribution
with covariance matrix $\boldsymbol{\Sigma}$ and precision matrix
$\mathbf{K} = \boldsymbol{\Sigma}^{-1}$. A Gaussian graphical model represents
conditional independences among the variables by zeros in the precision matrix.
For $i \neq j$, the pairwise Markov property states that

$$
K_{ij} = 0
\quad \Longleftrightarrow \quad
X_i \perp\!\!\!\perp X_j \mid \mathbf{X}_{-(i,j)} .
$$

Thus an absent edge corresponds to conditional independence given all remaining
variables in the model (Lauritzen, 1996).

### Partial-correlation identity

The edge weights in a GGM are usually reported as partial correlations rather
than raw precision-matrix entries. The partial correlation between variables
$i$ and $j$, conditional on all other variables, is

$$
\rho_{ij \cdot \mathrm{rest}}
= -\frac{K_{ij}}{\sqrt{K_{ii}K_{jj}}}.
$$

This transformation puts the conditional association on the familiar correlation
scale. A positive value indicates a positive conditional association, a negative
value indicates a negative conditional association, and zero corresponds to no
edge in the graph.

### The L1-penalized Gaussian log-likelihood objective

Let $\mathbf{S}$ be the sample covariance or correlation matrix. The graphical
lasso estimates the precision matrix by maximizing the penalized Gaussian
log-likelihood

$$
\hat{\mathbf{K}}
= \arg\max_{\mathbf{K} \succ 0}
\left[
\log\det(\mathbf{K})
- \operatorname{tr}(\mathbf{S}\mathbf{K})
- \lambda \sum_{i \neq j} |K_{ij}|
\right].
$$

The penalty parameter $\lambda \geq 0$ controls sparsity. Larger values shrink
more off-diagonal precision entries to zero, producing fewer edges; smaller
values retain more edges. This is the statistical source of the sparse network
estimated by the graphical lasso (Friedman, Hastie & Tibshirani, 2008).

### EBIC with gamma

The EBIC selects among candidate sparse graphs by balancing model fit against
model complexity. For a graph with $E$ non-zero edges estimated from $n$
observations and $p$ variables, a common form is

$$
\mathrm{EBIC}_{\gamma}
= -2\ell(\hat{\mathbf{K}})
+ E \log(n)
+ 4 E \gamma \log(p),
$$

where $\ell(\hat{\mathbf{K}})$ is the maximized Gaussian log-likelihood and
$\gamma$ controls the additional penalty for the size of the graph space (Foygel
& Drton, 2010). When $\gamma = 0$, EBIC reduces to the ordinary BIC penalty for
the number of edges. Larger values of $\gamma$ favour sparser networks.

## References

Borsboom, D., Deserno, M. K., Rhemtulla, M., Epskamp, S., Fried, E. I., McNally,
R. J., Robinaugh, D. J., Perugini, M., Dalege, J., Costantini, G., Isvoranu, A.-M.,
Wysocki, A. C., van Borkulo, C. D., van Bork, R., & Waldorp, L. J. (2021).
Network analysis of multivariate data in psychological science. *Nature Reviews
Methods Primers*, 1, Article 58.

Bringmann, L. F., Elmer, T., Epskamp, S., Krause, R. W., Schoch, D., Wichers, M.,
Wigman, J. T. W., & Snippe, E. (2019). What do centrality measures measure in
psychological networks? *Journal of Abnormal Psychology*, 128(8), 892-903.

Epskamp, S., Borsboom, D., & Fried, E. I. (2018). Estimating psychological
networks and their accuracy: A tutorial paper. *Behavior Research Methods*,
50(1), 195-212.

Epskamp, S., & Fried, E. I. (2018). A tutorial on regularized partial correlation
networks. *Psychological Methods*, 23(4), 617-634.

Foygel, R., & Drton, M. (2010). Extended Bayesian information criteria for
Gaussian graphical models. In *Advances in Neural Information Processing Systems*
(Vol. 23, pp. 604-612).

Friedman, J., Hastie, T., & Tibshirani, R. (2008). Sparse inverse covariance
estimation with the graphical lasso. *Biostatistics*, 9(3), 432-441.

Haslbeck, J. M. B., & Waldorp, L. J. (2018). How well do network models predict
observations? On the importance of predictability in network models. *Behavior
Research Methods*, 50(2), 853-861.

Lauritzen, S. L. (1996). *Graphical Models*. Oxford University Press.

Liu, H., Lafferty, J., & Wasserman, L. (2009). The nonparanormal: Semiparametric
estimation of high dimensional undirected graphs. *Journal of Machine Learning
Research*, 10, 2295-2328.

Saqr, M., Beck, E., & López-Pernas, S. (2024). Psychological networks: A modern
approach to the analysis of learning and complex learning processes. In M. Saqr
& S. López-Pernas (Eds.), *Learning Analytics Methods and Tutorials: A Practical
Guide Using R* (Chapter 19). Springer.
<https://lamethods.org/book1/chapters/ch19-psychological-networks/ch19-psych.html>
