---
title: "Gaussian graphical models: an extended tutorial"
author: "[Mohammed Saqr](https://saqr.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian graphical models: an extended tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 5)
library(psychnets)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
has_glasso <- requireNamespace("glasso", quietly = TRUE)
```

## What a psychological network represents

A psychological network is a statistical representation of relations among
measured variables. In a Gaussian graphical model the nodes are variables:
symptoms, questionnaire items, scale scores, behaviours, or other measured
constructs. The edges are partial correlations. An edge states that two variables
remain associated after conditioning on all other variables in the network, a
conditional dependence. The absence of an edge states that two variables are
conditionally independent given the rest, so any observed bivariate relation
between them is accounted for through other variables in the model.

This extended tutorial assumes the reader has met the model in the companion
vignette. It concentrates on interpretation and on the questions a careful
analysis must answer past the point estimate: how accurate the edge weights are,
how stable the centrality ordering is, and how the choice of estimation engine
affects the result. The `SRL_GPT` data set runs through the whole tutorial. It
contains 300 observations of five Motivated Strategies for Learning Questionnaire
construct scores: cognitive strategy use (`CSU`), intrinsic value (`IV`),
self-efficacy (`SE`), self-regulation (`SR`), and test anxiety (`TA`), each a
mean score on a 1 to 7 scale.

```{r}
head(SRL_GPT)
```

## Fitting and reading the network

`psychnet()` takes a numeric table and a `method` name and returns a fitted
`psychnet` object. Here `method = "glasso"` requests the
EBIC-regularized graphical lasso, which selects an L1 penalty by the extended
Bayesian information criterion and refits it to the certified optimum.

```{r}
net <- psychnet(SRL_GPT, method = "glasso")
net
```

The network has five nodes and ten edges at a selected penalty of `lambda` =
0.0086 and the default EBIC hyperparameter `gamma` = 0.5. The edge table is read
with `summary()`, which also reports the range and mean of the weights.

```{r}
summary(net)
```

Each row is a conditional association between two constructs. A positive weight
means that respondents higher on one construct tend to be higher on the other
after conditioning on the remaining constructs; a negative weight means the
opposite conditional relation. The learning and motivation constructs form a
positive cluster: `CSU` connects to `IV` (0.41), `SE` (0.38), and `SR` (0.35),
and `IV`, `SE`, and `SR` are positively linked among themselves. The one
substantial negative association is between `SR` and `TA` (-0.35), which reads as
lower test anxiety among respondents with higher self-regulation, holding the
other constructs constant. The three positive edges involving `TA` are small
(0.05 to 0.11), so their substantive interpretation is cautious. These weights are
conditional relations among construct scores and carry no causal claim.

`net_centralities()` summarizes how each node is positioned in the network,
returning a tidy data frame with one row per node and a column per measure. Node strength is the sum of the absolute edge weights at a node.
Expected influence is the signed sum, so positive and negative edges offset one
another.

```{r}
net_centralities(net)
```

`SR` has high strength (1.23) because it connects to several learning constructs
and carries the negative edge with `TA`. Its expected influence (0.53) is lower
than its strength because the negative edge offsets its positive connections.
Strength is usually the most interpretable and comparatively stable centrality
index in psychological networks. Betweenness and closeness are often unstable in
these networks and are available on request but are read with more caution.

`net_predict()` reports how well each node is predicted by its neighbours,
returning a tidy data frame with columns `node`, `type`,
`metric`, `predictability`, and `accuracy`. For a Gaussian graphical model the
value is the closed-form proportion of variance of each node explained by the rest
of the network, computed from the precision matrix.

```{r}
net_predict(net)
```

Predictability is an absolute complement to centrality: two nodes can have similar
centrality yet differ in how well the network explains them. The learning
constructs are highly predictable (`CSU` 0.82, `SR` 0.78, `IV` 0.77, `SE` 0.71),
whereas `TA` is much less predictable (0.14). That pattern fits the edge table:
test anxiety has one substantial edge and three weak ones.

```{r ggm-plot, eval = has_cograph}
cograph::splot(net, psych_styling = TRUE)
```

The graph is a visual summary of the edge table and is useful for orientation. The
numerical edge table remains the primary object for interpretation.

## Assessing accuracy of edge weights

A reported edge weight is one estimate from one sample. Before an edge is read as
substantive, its sampling variability is assessed. `net_boot()` resamples the
observations with replacement, re-estimates the network on each resample, and
summarizes the sampling distribution of every edge weight and
centrality. It returns an object of class `psychnet_bootstrap` whose edge table
carries the observed weight, the bootstrap mean, the lower and upper percentile
confidence limits, the inclusion proportion, and a `significant` flag set when the
percentile interval excludes zero. The following run uses 200 resamples for a
compact vignette; a full analysis uses more.

```{r}
set.seed(1)
bs <- net_boot(SRL_GPT, method = "glasso", n_boot = 200, cores = 1)
bs
```

The printed summary reports that seven of the ten edges are significant at the
95% level. The full edge table gives the intervals.

```{r}
summary(bs)
```

A percentile confidence interval is the range of edge values consistent with the
data at the stated level. An interval that excludes zero indicates a non-zero
edge. The strong edges of the positive cluster have narrow intervals well away
from zero: `CSU`-`IV` runs from 0.32 to 0.49, and the negative `SR`-`TA` edge runs
from -0.43 to -0.24. The three weak positive `TA` edges have intervals whose lower
limit is zero, so they are not significant, and their inclusion proportions are
below one (`CSU`-`TA` at 0.72, `SE`-`TA` at 0.96, `IV`-`TA` at 0.97), meaning the
edge was set to zero in a share of the resamples. The bootstrap therefore gives
the same reading as the point estimate but with the added evidence that the weak
`TA` edges are uncertain.

## Assessing stability of centrality

Centrality indices are functions of many estimated edges, so their ordering also
needs a stability check. `net_stability()` drops random subsets of cases,
re-estimates the network on each subset, and correlates the subset centralities
with the full-sample centralities. It returns an object of
class `psychnet_stability` carrying the correlation-stability (CS) coefficient per
measure and a tidy table of the mean correlation at each drop proportion. The CS
coefficient is the largest proportion of cases that can be dropped while the rank
correlation with the full sample stays at or above a threshold (default 0.7) with
a stated certainty (default 0.95).

```{r}
set.seed(1)
cs <- net_stability(SRL_GPT, method = "glasso",
                    drop_prop = c(0.3, 0.5, 0.7), iter = 50)
cs
```

The CS coefficient is 0.50 for strength and 0.70 for expected influence. The
strength ordering stays correlated at 0.7 or above with the full-sample ordering,
with 95% certainty, when up to 50% of the cases are dropped; the expected-influence
ordering holds up to 70%. A common guideline treats a CS coefficient of at least
0.5 as adequate and 0.25 as a minimum, so both measures are stable enough to
interpret here. The supporting table shows how the mean correlation declines as
more cases are dropped.

```{r}
cs$table
```

For strength the mean correlation falls from 0.97 at a 30% drop to 0.89 at a 70%
drop; for expected influence it falls from 0.998 to 0.96 over the same range. The
higher correlations for expected influence match its larger CS coefficient.

## Native and external estimation engines

By default (`native = TRUE`) the graphical lasso is estimated with the package's
own solver, written entirely in base R. This is a clean-room implementation: the
algorithm is reconstructed from first principles and reproduces the estimate
without wrapping an external library. It depends only on base R, so it installs
and runs anywhere with no compilation; it is readable R code that can be inspected
end to end; and each fit is graded against the convex objective it optimizes, so
its correctness is read from the certificate.

Setting `native = FALSE` delegates each penalized solve to the compiled `glasso`
Fortran routine, the same code underlying `qgraph::EBICglasso()`, which is faster
on large problems. The optional `glasso` package supplies it.

```{r fortran-fit, eval = has_glasso}
net_fortran <- psychnet(SRL_GPT, method = "glasso", native = FALSE)
summary(net_fortran)
```

The two solvers estimate the same model and return the same network: the edge
weights agree with the native fit to about five decimal places (`CSU`-`IV` is
0.4117 under both). The certificate reads the difference between them.

```{r fortran-certificate, eval = has_glasso}
certificate(net_fortran)
```

The residual here is 8.2e-05, above the default tolerance of 1e-6, so `certified`
is `FALSE`. The Fortran routine converges to its own looser tolerance (near
1e-4), and the certificate reports that looser tolerance; the base-R refit reaches
a tighter one. The native fit certifies at 2.2e-10, so both solvers recover the same
network while the base-R solver returns it to a tighter numerical optimum. A study
can use the native solver and read the external one as an independent check.

## Non-normal and ordinal data

Gaussian graphical models are Gaussian, yet psychological data are often skewed,
bounded, and measured on ordinal response scales. Scale means such as the five
`SRL_GPT` construct scores are commonly treated as approximately continuous, which
is an approximation of the questionnaire response format.

For ordinal item-level data, `cor_method = "auto"` bases the correlation input on
polychoric or polyserial associations when appropriate. This choice is most
relevant when the variables are individual Likert items or mixed
ordinal-continuous measures.

For continuous variables with pronounced monotone departures from normality,
`method = "huge"` estimates a nonparanormal graphical model. The nonparanormal
assumes the variables become jointly Gaussian after unknown monotone
transformations of their marginal distributions, then applies the identical EBIC
graphical lasso to the transformed correlation matrix. It retains the
conditional-dependence interpretation of the Gaussian graphical model while
relaxing strict multivariate normality.

```{r}
huge_net <- psychnet(SRL_GPT, method = "huge")
certificate(huge_net)
```

The nonparanormal fit certifies at 2.5e-10, at the order of machine precision, so
it reached the optimum of its convex objective on the transformed correlation.

## Mathematical foundations

This section gives the statistical definitions behind the tutorial. It can be
skipped without losing the worked example.

### Precision matrix and the pairwise Markov property

Let $\mathbf{X} = (X_1, \ldots, X_p)$ follow a multivariate normal distribution
with covariance matrix $\boldsymbol{\Sigma}$ and precision matrix
$\mathbf{K} = \boldsymbol{\Sigma}^{-1}$. The model represents conditional
independences by zeros in the precision matrix. For $i \neq j$,

$$
K_{ij} = 0
\quad \Longleftrightarrow \quad
X_i \perp\!\!\!\perp X_j \mid \mathbf{X}_{-(i,j)} ,
$$

so an absent edge is a conditional independence given all remaining variables.

### Partial-correlation identity

The edge weights are reported as partial correlations. The partial correlation
between variables $i$ and $j$, conditional on all other variables, is

$$
\rho_{ij \cdot \mathrm{rest}}
= -\frac{K_{ij}}{\sqrt{K_{ii}K_{jj}}} ,
$$

which puts the conditional association on the correlation scale: a positive value
is a positive conditional association, a negative value a negative one, and zero
is no edge.

### The L1-penalized Gaussian log-likelihood

Let $\mathbf{S}$ be the sample correlation matrix. The graphical lasso estimates
the precision matrix by maximizing the penalized Gaussian log-likelihood,

$$
\hat{\mathbf{K}}
= \arg\max_{\mathbf{K} \succ 0}
\left[
\log\det(\mathbf{K})
- \operatorname{tr}(\mathbf{S}\mathbf{K})
- \lambda \sum_{i \neq j} |K_{ij}|
\right] .
$$

The penalty $\lambda \geq 0$ controls sparsity: larger values shrink more
off-diagonal entries to zero and yield fewer edges. The objective is strictly
concave in $\mathbf{K}$, so its maximizer is unique.

### EBIC with gamma

The extended Bayesian information criterion selects among candidate sparse graphs
by balancing fit against complexity. For a graph with $E$ non-zero edges from $n$
observations and $p$ variables,

$$
\mathrm{EBIC}_{\gamma}
= -2\ell(\hat{\mathbf{K}})
+ E \log(n)
+ 4 E \gamma \log(p) ,
$$

where $\ell(\hat{\mathbf{K}})$ is the maximized Gaussian log-likelihood and
$\gamma$ controls the additional penalty for the size of the graph space. When
$\gamma = 0$ the criterion reduces to ordinary BIC; larger $\gamma$ favours
sparser networks.

### Bootstrap and case-dropping stability

The nonparametric bootstrap draws resamples of the observations with replacement,
re-estimates the network on each, and forms the percentile confidence interval of
each edge weight and centrality from the resampling distribution. The
case-dropping procedure removes random fractions of the cases, re-estimates on
each subset, and correlates the subset centralities with the full-sample
centralities. The correlation-stability coefficient is the largest drop proportion
at which that correlation stays at or above a threshold with a stated certainty.
Both procedures re-estimate through the same `psychnet()` path, so they apply to
any estimator with no method-specific code.
