---
title: "Gaussian graphical models: estimation and interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian graphical models: estimation and interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
library(psychnets)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## What a Gaussian graphical model is

A Gaussian graphical model represents the conditional dependence structure of
a set of continuous variables. Each node is a measured variable. Each edge is a
partial correlation between two variables after conditioning on every other
variable in the model. The graph is undirected because the partial correlation
between variables $i$ and $j$ has no direction.

The practical question is whether two variables remain related above and beyond
what the other variables can explain. Consider a network containing motivation,
academic achievement, prior achievement, study effort, self-efficacy, and test
anxiety. An edge between motivation and academic achievement means that their
association remains after accounting for the other four variables. The edge is
the part of their relationship that cannot be explained by prior achievement,
study effort, self-efficacy, or test anxiety within the fitted model.

If the motivation-achievement edge is absent, the model did not retain a unique
association between them after accounting for those other variables. Their
ordinary correlation may still be positive. That correlation might be explained
by motivated students studying more, having greater self-efficacy, or differing
in prior achievement. The network helps separate this shared association from
the relation that remains after the other measured variables are considered.

The estimation procedure also filters negligible connections that may reflect
sampling noise. A missing edge therefore has a practical reading: the data and
the selected model did not provide enough unique association to retain that
connection. It does not mean that the variables are unrelated in every context.
Edges describe adjusted associations and do not by themselves show that one
variable causes another.

## The data

The worked example uses `SRL_GPT`, a data frame supplied with `psychnets`. It
contains 300 observations on five construct scores from a self-regulated
learning questionnaire. Cognitive strategy use (`CSU`), intrinsic value (`IV`),
self-efficacy (`SE`), self-regulation (`SR`), and test anxiety (`TA`) are means
of item responses recorded on a 1 to 7 scale.

```{r data-preview}
head(SRL_GPT)
```

Rows are observations and columns are network nodes. The analysis treats the
five scale scores as continuous variables. This treatment is most defensible
when each score aggregates several items and has enough distinct values to
approximate a continuous distribution. Item-level ordinal data may require an
ordinal correlation estimator, discussed under sensitivity analysis.

Before fitting a network, a scientific analysis should also examine missingness,
univariate distributions, unusual observations, and the effective sample size.
The current data are complete. The default estimator therefore uses the same 300
observations for every pairwise correlation.

## Fitting the network with `psychnet()`

`psychnet()` estimates a network from a numeric table and returns a fitted
`psychnet` object. The argument `method = "glasso"` requests an
EBIC-regularized Gaussian graphical model. The default settings use Pearson
correlations, pairwise-complete handling of missing values, $\gamma = 0.5$, and
the package's native graphical-lasso solver.

```{r fit-network}
net <- psychnet(data = SRL_GPT, method = "glasso")
net
```

The printed object reports 5 nodes and 10 retained edges. The selected penalty
is $\lambda = 0.00861$ at $\gamma = 0.5$. All possible pairs are retained in
this small example, although several estimated edges are weak. The selected
penalty and the observed data determine the degree of sparsity in each fitted
network.

## Inspecting edges with `summary()`

`summary()` prints the fitted model and returns its edge table invisibly. The
table contains the columns `from`, `to`, and `weight`. Each row identifies a
retained pair of nodes and its estimated partial correlation.

```{r summarize-network}
summary(net)
```

The edge weights range from -0.350 to 0.412, with a mean of 0.174 across the 10
retained edges. The positive `CSU`-`IV` edge is the largest positive conditional
association ($r = 0.412$). Cognitive strategy use and intrinsic value remain
positively associated after adjustment for self-efficacy, self-regulation, and
test anxiety.

The `SR`-`TA` edge is negative ($r = -0.350$). Higher self-regulation is
associated with lower test anxiety among observations with comparable values on
the other three constructs. The edges from `TA` to `CSU` ($r = 0.049$), `SE`
($r = 0.099$), and `IV` ($r = 0.111$) are small. Their magnitudes support only
cautious substantive interpretation. None of these conditional associations
identifies a causal effect.

## Checking numerical optimality with `certificate()`

`certificate()` reports how closely a fitted estimator satisfies the defining
conditions of its optimization problem. For a graphical-lasso fit, the returned
data frame has the columns `method`, `certificate`, `kind`, and `certified`.
The `certificate` value is the maximum violation of the Karush-Kuhn-Tucker (KKT)
stationarity conditions. The `kind` value is `"kkt"`, and `certified` records
whether the residual is at or below the requested tolerance of $10^{-6}$ by
default.

```{r optimality-certificate}
certificate(net)
```

The KKT residual is $2.21 \times 10^{-10}$ and `certified` is `TRUE`. This
residual is close to machine precision. It indicates that the estimated
precision matrix satisfies the first-order conditions of the selected convex
objective to numerical precision. The certificate assesses numerical
optimization. It does not assess the Gaussian assumption, the substantive
validity of the variables, sampling uncertainty, or causal identification.

## Describing node position with `net_centralities()`

`net_centralities()` computes node-level summaries from the fitted weighted
graph. Its default return is a tidy data frame with the columns `node`,
`strength`, and `expected_influence`. Strength is the sum of the absolute edge
weights incident on a node. Expected influence is the sum of the signed edge
weights, so negative and positive relations can offset one another.

```{r node-centrality}
net_centralities(net)
```

Self-regulation has the largest strength (1.231), followed by cognitive strategy
use (1.198). Self-regulation combines several positive connections with the
negative connection to test anxiety. Its expected influence is therefore 0.532,
which is lower than its strength. Cognitive strategy use has only positive
estimated edges, so its strength and expected influence are both 1.198.

Test anxiety has strength 0.608 and expected influence -0.091. Its negative
edge with self-regulation offsets its three small positive edges. This example
shows why the two indices answer different questions. Strength quantifies total
absolute connection, whereas expected influence retains the direction of each
relation. Centrality is a descriptive property of the estimated graph. A high
value does not identify a variable as a cause or an intervention target.

## Quantifying node predictability with `net_predict()`

`net_predict()` reports how well every node is predicted by the other nodes in a
fitted network. For a Gaussian graphical model, it computes the proportion of
variance explained, $R^2$, directly from the precision and covariance matrices.
The returned data frame has the columns `node`, `type`, `metric`,
`predictability`, and `accuracy`. The Gaussian rows use `type = "gaussian"`,
`metric = "R2"`, and `accuracy = NA` because classification accuracy does not
apply to continuous nodes.

```{r node-predictability}
net_predict(net)
```

Cognitive strategy use has the largest predictability, with $R^2 = 0.822$.
Self-regulation ($R^2 = 0.779$), intrinsic value ($R^2 = 0.769$), and
self-efficacy ($R^2 = 0.714$) are also strongly explained by their remaining
nodes in this fitted model. Test anxiety has $R^2 = 0.142$, so the other four
constructs account for a much smaller proportion of its variance.

Predictability complements centrality. Centrality summarizes how a node is
connected, while predictability quantifies how much of its variance is accounted
for by the other nodes. Predictability remains an in-sample model quantity in
this analysis. It should not be read as out-of-sample predictive performance
without a separate validation design.

## Visualizing the fitted network with `cograph::splot()`

`cograph::splot()` draws the fitted object as a network when the optional
`cograph` package is available. The argument `psych_styling = TRUE` uses green
edges for positive weights, red edges for negative weights, and node rings for
the stored predictability values.

```{r plot-network, eval = has_cograph}
cograph::splot(net, psych_styling = TRUE)
```

The figure is a compact representation of the edge table. Edge colour encodes
sign and edge width encodes magnitude. Node position is determined by a layout
algorithm and has no direct statistical scale. Visual distance between two nodes
must therefore not be interpreted as an estimated association. Numerical
interpretation should return to the edge table and node-level summaries.

## Sensitivity to the correlation estimator

Analytic choices can affect weak edges. A sensitivity analysis changes one
choice at a time and compares the resulting model with the primary fit.
`psychnet()` accepts `cor_method = "spearman"` for a rank-based analysis.
Spearman correlations reduce sensitivity to monotone transformations and
extreme values, although they answer a rank-association version of the research
question.

```{r fit-spearman}
spearman_net <- psychnet(data = SRL_GPT, method = "glasso", cor_method = "spearman")
spearman_net
```

```{r summarize-spearman}
summary(spearman_net)
```

The Spearman analysis retains 9 edges and selects $\lambda = 0.01882$. The main
pattern remains: the four learning constructs are positively connected and the
`SR`-`TA` edge is negative. The weak `CSU`-`TA` edge is absent, and the
`SR`-`TA` estimate changes from -0.350 to -0.237. The persistence of the main
pattern provides evidence that its qualitative interpretation is not specific to
Pearson correlations. The changes in weaker edges indicate sensitivity in that
part of the graph.

For item-level ordinal variables, `cor_method = "auto"` can estimate
polychoric or polyserial associations where appropriate. The measurement scale
should determine this choice before the network results are examined.

## Applying a substantive edge threshold

`psychnet()` can apply an absolute post-estimation threshold through its
`threshold` argument. A threshold is a reporting or sparsification choice. It
does not change the penalty selected by EBIC, and it should be justified before
examining the final graph.

```{r fit-threshold}
threshold_net <- psychnet(data = SRL_GPT, method = "glasso", threshold = 0.1)
threshold_net
```

```{r summarize-threshold}
summary(threshold_net)
```

At a threshold of 0.10, the graph contains 8 edges. The `CSU`-`TA` estimate
(0.049) and the `SE`-`TA` estimate (0.099) are removed. The selected penalty and
KKT residual are unchanged because thresholding occurs after estimation. A
threshold can make a graph easier to read, but it must not be presented as a
model-selection result from EBIC.

## A nonparanormal sensitivity model

The Gaussian graphical model assumes joint normality for its exact probabilistic
interpretation. `psychnet()` fits a rank-based nonparanormal model with
`method = "huge"`. This method estimates monotone marginal transformations and
then applies the same EBIC graphical-lasso procedure to the transformed
correlation matrix. Its purpose is to relax strict marginal normality while
retaining a Gaussian conditional-dependence model on the transformed scale.

```{r fit-nonparanormal}
npn_net <- psychnet(data = SRL_GPT, method = "huge")
npn_net
```

```{r summarize-nonparanormal}
summary(npn_net)
```

The nonparanormal model retains all 10 edges. Its weights range from -0.343 to
0.426. The positive cluster among the four learning constructs and the negative
`SR`-`TA` edge remain. The weakest `CSU`-`TA` edge decreases from 0.049 to
0.024, which again marks this edge as sensitive to the correlation model.

```{r certify-nonparanormal}
certificate(npn_net)
```

The KKT residual is $2.47 \times 10^{-10}$ and `certified` is `TRUE`. The
optimization therefore satisfies its first-order conditions to numerical
precision on the transformed correlation matrix. This result checks the solver,
while the scientific adequacy of the nonparanormal model remains a modeling
judgment.

## Reporting the analysis

A complete report should identify the variables and their measurement scale,
the sample size, the correlation estimator, the missing-data procedure, the
network estimator, the EBIC value of $\gamma$, and any post-estimation
threshold. It should report the edge weights numerically and distinguish
conditional association from causation. Centrality and predictability should be
defined before their values are interpreted.

The present analysis used Pearson correlations and an EBIC graphical lasso with
$\gamma = 0.5$ on 300 complete observations. The selected model had 5 nodes,
10 edges, and $\lambda = 0.00861$. Its KKT residual was
$2.21 \times 10^{-10}$. The largest positive edge joined cognitive strategy use
and intrinsic value ($r = 0.412$), and the largest negative edge joined
self-regulation and test anxiety ($r = -0.350$). Rank-based and nonparanormal
sensitivity models preserved the main qualitative pattern but changed some weak
edges.

## Why the model removes weak connections

The graphical lasso filters small connections so that the fitted network focuses
on associations with enough unique information to be retained. This filtering
is useful because a sample can contain many negligible correlations produced by
ordinary sampling variation. Keeping every one of them would make the network
harder to read and could give weak, noisy connections more attention than their
magnitudes justify.

The amount of filtering is selected from the data. `psychnets` fits networks
across a sequence of penalty values and uses the extended Bayesian information
criterion (EBIC) to balance fit with the number of retained edges. The
hyperparameter $\gamma$ controls how strongly EBIC favours a simpler network.
Larger values generally retain fewer edges.

This procedure explains how the graph was simplified. It does not make the
retained edges causal, and it does not show that every removed edge is exactly
zero in the population. Weak edges may change across samples or analytic
choices, which is why the Spearman and nonparanormal examples above are useful.

## Mathematical foundations

This section states the model, estimator, selection criterion, and numerical
diagnostic precisely. The worked example can be understood without these
derivations.

### Precision matrix and conditional independence

Let $\mathbf{X} = (X_1, \ldots, X_p)^\mathsf{T}$ follow a multivariate normal
distribution with mean vector $\boldsymbol{\mu}$ and positive-definite
covariance matrix $\boldsymbol{\Sigma}$. The precision matrix is
$\mathbf{K} = \boldsymbol{\Sigma}^{-1}$. For distinct nodes $i$ and $j$,

$$
K_{ij} = 0
\quad \Longleftrightarrow \quad
X_i \perp\!\!\!\perp X_j \mid
\mathbf{X}_{\setminus\{i,j\}}.
$$

Zeros in the precision matrix therefore define the missing edges of the
population graph. The partial correlation associated with a retained edge is

$$
\rho_{ij \cdot \mathrm{rest}}
= -\frac{K_{ij}}{\sqrt{K_{ii}K_{jj}}}.
$$

This standardization places conditional associations on the correlation scale,
from -1 to 1.

### Penalized likelihood

Let $\mathbf{S}$ be the sample covariance or correlation matrix. The graphical
lasso estimates a positive-definite precision matrix by minimizing

$$
\widehat{\mathbf{K}}
= \underset{\mathbf{K} \succ 0}{\operatorname{argmin}}
\left\{
-\log\det(\mathbf{K})
+ \operatorname{tr}(\mathbf{S}\mathbf{K})
+ \lambda \sum_{i \ne j}|K_{ij}|
\right\}.
$$

The tuning parameter $\lambda \geq 0$ controls the amount of regularization.
Larger values impose more shrinkage on off-diagonal precision entries and
usually produce fewer edges. The objective is strictly convex over the cone of
positive-definite matrices, so its minimizer is unique.

### EBIC model selection

For each candidate value of $\lambda$, the fitted graph has a Gaussian
log-likelihood $\ell(\widehat{\mathbf{K}})$ and $E$ retained edges. The EBIC is

$$
\operatorname{EBIC}_{\gamma}
= -2\ell(\widehat{\mathbf{K}})
+ E\log(n)
+ 4E\gamma\log(p),
$$

where $n$ is the sample size, $p$ is the number of nodes, and $\gamma$ controls
the additional graph-complexity penalty. `psychnets` evaluates a logarithmically
spaced path of candidate penalties and selects the value with the smallest EBIC.
The selected penalty is then refitted at a tighter numerical tolerance.

### KKT optimality conditions

Write $\mathbf{W} = \widehat{\mathbf{K}}^{-1}$. At the optimum, the diagonal
conditions satisfy $W_{ii} = S_{ii}$. For off-diagonal entries,

$$
W_{ij} - S_{ij}
= \lambda\,\operatorname{sign}(\widehat{K}_{ij})
\quad \text{when } \widehat{K}_{ij} \ne 0,
$$

and

$$
|W_{ij} - S_{ij}| \leq \lambda
\quad \text{when } \widehat{K}_{ij} = 0.
$$

`certificate()` returns the maximum violation of these conditions. A residual
near zero indicates that the numerical solution satisfies the defining
first-order conditions to numerical precision.

### Predictability from the precision matrix

For node $j$, the conditional residual variance is $1/K_{jj}$. If $S_{jj}$ is
the marginal variance, the proportion of variance explained by the remaining
nodes is

$$
R_j^2 = 1 - \frac{1}{K_{jj}S_{jj}}.
$$

`net_predict()` evaluates this identity for each node. The calculation requires
the fitted precision and covariance matrices, so raw observations are not needed
after a Gaussian graphical model has been fitted.
