---
title: "Group networks from event logs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group networks from event logs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.4)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
library(psychnets)
```

```{r simulate, include = FALSE}
# A small simulated learning-process log: students from two cohorts each emit a
# sequence of self-regulation actions. Both cohorts have a tightly co-occurring
# cluster of three states -- but a DIFFERENT cluster: in "focused" it is
# plan/monitor/reflect, in "scattered" it is reflect/help/offtask. The two
# networks therefore carry a similar amount of structure in different places. The
# log is the kind of trace a learning-management system produces, one row per
# action.
set.seed(1)
states <- c("plan", "monitor", "reflect", "help", "offtask")
make_cohort <- function(name, cluster, n_students) {
  f <- stats::rnorm(n_students)                       # latent regulation tendency
  lambda <- rep(0.1, 5)                               # weak by default ...
  lambda[cluster] <- 1.0                              # ... strong inside the cluster
  mu <- outer(f, lambda) + matrix(stats::rnorm(n_students * 5, 0, .4),
                                  n_students, 5)
  counts <- matrix(pmax(0, round(6 + 3 * mu)),        # non-negative action counts
                   n_students, 5)                     # (pmax drops the dims)
  per_student <- lapply(seq_len(n_students), function(i) {
    act <- rep(states, counts[i, ])
    if (!length(act)) act <- sample(states, 1)
    data.frame(student = sprintf("%s%03d", name, i), cohort = name,
               action = act, stringsAsFactors = FALSE)
  })
  do.call(rbind, per_student)
}
events <- rbind(make_cohort("focused",   1:3, 100),
                make_cohort("scattered", 3:5, 100))
events <- events[sample(nrow(events)), ]              # shuffle into log order
rownames(events) <- NULL
```

Much psychological and learning-process data arrives as an **event log** — one
row per action taken by a person over time — rather than as a tidy
person-by-variable table. `psychnet()` reads such a log directly: name the
`actor` and `action` columns and it builds each person's profile internally,
with no manual reshaping. Adding `group` estimates a **separate network for each
group** and returns a collection you can plot, bootstrap, and compare with the
same verbs you would use on a single network.

## An event log

The example is a simulated trace of `r length(unique(events$student))` students
from two cohorts, each emitting one of five self-regulation actions per row.

```{r}
head(events)
```

## One network per group

Name the actor and action columns and pass `group`. `psychnet()` reduces each
student to their action profile, splits by cohort, and estimates a graphical
lasso network per group. The result is a `psychnet_group`: a named collection of
networks. The grouping column is never itself a node.

```{r}
nets <- psychnet(events, actor = "student", action = "action", group = "cohort")
nets
```

It plots as a grid, one panel per group, with a shared node layout so the same
state sits in the same place in both panels.

```{r grid, eval = has_cograph, fig.width = 8, fig.height = 4.2}
cograph::splot(nets, psych_styling = TRUE)
```

## Centrality, side by side

Every framework verb detects the group object and returns one result per group;
there is no per-group looping for the analyst to write. `as.data.frame()` stacks
the per-group tables under a `group` column.

```{r}
net_centralities(nets) |> as.data.frame()
```

## Predictability

```{r}
net_predict(nets) |> as.data.frame()
```

## Bootstrapped edge stability

`net_boot()` resamples students within each group, re-estimates, and returns a
percentile confidence interval and a `significant` flag (interval excludes zero)
for every edge. Ask for the significant edges with an argument rather than
subsetting the result by hand.

```{r}
boot <- net_boot(nets, n_boot = 150)
as.data.frame(boot, significant = TRUE)
```

## Comparing the two networks

`net_compare()` runs a permutation network-comparison test between two groups:
global strength (M), overall structure (S), and per-edge differences (E).

```{r}
net_compare(nets, c("focused", "scattered"), iter = 250)
```

The two cohorts were built to carry a **similar amount** of regulation structure
in **different places** (a strong plan/monitor/reflect cluster versus a strong
reflect/help/offtask one). That is what the test recovers: global strength (M) is
comparable, while the network structure (S) differs.

## How it works

An event log has no person-by-variable matrix to estimate a network on, so
`psychnet()` builds one. Each occasion (here, each student) becomes a row, each
distinct action a column, and the cell holds that action's count — the same
"frequency" representation `event_frequencies()` returns. The network is the
graphical lasso of that matrix.

`group` simply splits the log before this reduction and estimates one network per
level. Because each group is reduced to an ordinary numeric design matrix, the
group object resamples and compares exactly like cross-sectional data — which is
why `net_boot()` and `net_compare()` work on it with no special handling.

A `time` column, if present, is used only to split each actor into sessions from
gaps between events (`compute_sessions`); there is no temporal or lagged model
here. When a student contributes several occasions, the actor clustering is
removed by person-centering (`standardize = TRUE`, the default), or — with
`standardize = FALSE` — decomposed into within- and between-actor networks.
