---
title: "Regulation networks from event data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regulation networks from event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_tna     <- requireNamespace("tna", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 5.5, warning = FALSE, message = FALSE,
                      eval = has_tna)
library(psychnets)
if (has_tna) {
  data("group_regulation_long", package = "tna")
  gr <- as.data.frame(group_regulation_long)
}
```

## About psychnets

`psychnets` is the psychometric-network engine of **Dynalytics**, a family of
tools for analysing the *process* and *event* data that learning environments and
log files produce. It builds a psychometric network **directly from an event
log** — you do not have to reshape the log into a wide person-by-item table
first. It handles two things for you:

- **It counts the actions (computes frequencies).** Each unit of observation
  becomes one row, and each action becomes a column holding how many times that
  action occurred. Those counts are the variables the network relates.
- **It respects nesting.** When the same person is observed more than once, their
  repeated observations are *nested* inside them. `psychnets` keeps the two kinds
  of variation apart — how states move together *within* a person versus how
  people *differ* on average — instead of mixing them into one misleading number.

### The three words you need

- **Action** — the thing that happened (here, a regulation state such as *plan*
  or *monitor*). Actions become the **nodes** of the network.
- **Actor** — *who* did it. Here the actor is the **student** (`Actor`).
- **Occasion** — one row of counts. By default an actor gives **one** occasion
  (all their actions summed). If you split each actor into **sessions**, one actor
  gives **several** occasions — and that is when de-clustering matters (below).

This vignette uses the `group_regulation_long` event log from the **tna**
package.

```{r no-tna, eval = !has_tna, echo = FALSE, results = "asis"}
cat("> The **tna** package is not installed, so the chunks below are not",
    "evaluated. Install it with `install.packages(\"tna\")` to run this",
    "vignette.\n")
```

## The data: a regulation event log

Each row is a single **action** by a **student** (`Actor`), at a **time**
(`Time`), in a collaboration **group** (`Group`), with the student's achievement
level (`Achiever`). This is the raw shape a learning platform records.

```{r data}
head(gr)
```

## Part 1 — one network from students' regulation profiles

Give `psychnet()` the actor column and it does the rest: it counts each
**student's** nine regulation actions into one profile, then estimates the
network of how those states co-occur across students. The result carries its own
correctness certificate — the tiny KKT residual — so you can trust it is the
optimal network for this data.

```{r plain}
net <- psychnet(gr, actor = "Actor")
net
```

```{r plain-plot, eval = has_tna && has_cograph}
cograph::splot(net, psych_styling = TRUE)
```

*In plain terms:* every student is summarised by how often they did each of the
nine actions, and the network shows which actions tend to rise and fall together
across students.

## Part 2 — one network per group

Add `group = "Achiever"` and `psychnet()` estimates a **separate network for each
level** — here, high- versus low-achieving students — and returns them as one
object that plots as a grid and is understood by every analysis verb.

```{r group}
byach <- psychnet(gr, actor = "Actor", group = "Achiever")
byach
```

```{r group-grid, eval = has_tna && has_cograph, fig.width = 11, fig.height = 5.5}
cograph::splot(byach, psych_styling = TRUE)
```

The framework verbs notice the group object and return one result per group —
you never loop over the groups yourself.

```{r group-cent}
net_centralities(byach)
```

## Part 3 — sessions, and why nesting needs de-clustering

So far each student gave **one** occasion. But a student's log is really several
short **working sessions**. If we treat a pause longer than five minutes as the
start of a new session (`time_threshold = 300` seconds), each student now
contributes **several** occasions — and those occasions are **nested** inside the
student. (With this data the sessions only appear at a short gap like this: each
student's raw log is one continuous episode, so the default 15-minute gap never
splits it.)

Why does that matter? Because two different questions get tangled together:

- **Within a student:** across *their own* sessions, which actions move together?
- **Between students:** do some students simply do more of everything than others?

If you ignore the nesting, these mix and the network can be misleading.
`psychnets` separates them.

### The de-clustered network (default)

By default (`standardize = TRUE`) `psychnet()` **removes the between-student
differences** — it centres each student's sessions on that student's own average
— and returns the single *within-student* network: how regulation states move
together inside a person's work, with overall student differences taken out.

```{r declustered}
net_within <- psychnet(gr, actor = "Actor", time = "Time", time_threshold = 300)
net_within
```

```{r declustered-plot, eval = has_tna && has_cograph}
cograph::splot(net_within, psych_styling = TRUE)
```

### Keeping both levels

Set `standardize = FALSE` to keep **both** networks: the **within-student**
network (co-movement across a person's sessions) and the **between-student**
network (how students' average profiles differ).

```{r within-between}
wb <- psychnet(gr, actor = "Actor", time = "Time", time_threshold = 300,
               standardize = FALSE)
wb
```

```{r wb-plots, eval = has_tna && has_cograph, fig.width = 11, fig.height = 5.5}
cograph::splot(wb$within,  psych_styling = TRUE)
cograph::splot(wb$between, psych_styling = TRUE)
```

Their centralities, each a single tidy call:

```{r wb-cent}
net_centralities(wb$within)
net_centralities(wb$between)
```

*In plain terms:* the **within** network answers "when this student is working,
which regulation states go hand in hand?"; the **between** network answers "how
do students differ from one another overall?". De-clustering is simply not
letting those two answers contaminate each other.

## Recap

- **Actor** = the student; **action** = the regulation state; **occasion** = one
  row of counts.
- With no sessions, one student = one occasion → **Part 1**, a single network;
  `group = "Achiever"` splits it into one network per group → **Part 2**.
- Split a student into **sessions** and they give many occasions nested inside
  them → **Part 3**, where de-clustering separates *within*-student co-movement
  from *between*-student differences (`standardize = TRUE` keeps only within;
  `FALSE` keeps both).
