---
title: "Edge-weight stability and split-half reliability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Edge-weight stability and split-half reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5.5, dpi = 100)
library(psychnets)
```

`net_stability()` tells you whether centrality *rankings* survive case-dropping.
This vignette covers the two complementary robustness checks for the *edge
structure* itself, both estimator-agnostic and base-R only:

- **`casedrop_reliability()`** — the edge-weight CS-coefficient: how many cases
  can be dropped before the whole edge-weight vector stops correlating with the
  full-sample one.
- **`network_reliability()`** — split-half reliability: split the sample, estimate
  a network on each half, and measure how alike the two edge structures are.

Each verb takes raw `data` and **returns a tidy `data.frame` you print
directly** — one call, no table assembly, no accessors.

We use the bundled `SRL_Claude` self-regulated-learning data (300 respondents,
five MSLQ subscales).

# Edge-weight case-dropping stability

One line in, a tidy table out — one row per metric per drop proportion, with the
CS-coefficient on the header line. No arguments needed:

```{r casedrop}
cd <- casedrop_reliability(SRL_Claude)
cd
```

`plot()` shows all four metrics against the drop proportion, with ±1 SD bands;
the correlation panel marks the acceptance threshold and the CS-coefficient.

```{r casedrop-plot, fig.width = 8, fig.height = 6.5}
plot(cd)
```

A CS-coefficient of 0.5 or higher is the usual rule of thumb for an edge
structure that can be interpreted with confidence.

# Split-half reliability

Again, one line in, a tidy per-metric table out:

```{r reliability}
rel <- network_reliability(SRL_Claude)
rel
```

`plot()` shows the distribution of each between-halves metric across the
split-half iterations, with the observed mean marked.

```{r reliability-plot, fig.width = 8, fig.height = 6.5}
plot(rel)
```

# Any estimator, any grouping

Both verbs route every refit through `psychnet(method = )`, so they work for any
estimator — swap in partial correlations, for instance:

```{r pcor}
casedrop_reliability(SRL_Claude, method = "pcor")
```

And both accept a `psychnet_group` (built with `psychnet(..., group = )`),
returning one result per level.

# Summary

| Verb | Question | Returns | Plot |
|---|---|---|---|
| `casedrop_reliability()` | How robust is the edge structure to dropping cases? | tidy df (metric × drop_prop) | four-metric curves + CS |
| `network_reliability()` | How reproducible is the edge structure across split-halves? | tidy df (one row per metric) | per-metric histograms |

Together with `net_stability()` (centrality rankings) and `net_boot()` (edge
accuracy intervals), these complete the bootnet robustness toolkit in base R.
