---
title: "Relative-importance networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Relative-importance networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

```{r}
library(psychnet)
```

A **relative-importance network** (`method = "relimp"`) is **directed**. For each
node taken as an outcome, it regresses it on all the others and decomposes that
regression's R-squared into each predictor's share using the **LMG / Shapley**
method (averaging the predictor's contribution over every ordering). The edge
`A -> B` is the share of B's variance explained by A. Unlike a GGM, the weights
sum meaningfully: a node's incoming edges add up to its full-model R-squared.

The Shapley decomposition enumerates predictor subsets, so it is meant for a
modest number of nodes. We use the five construct scores in `SRL_GPT`.

```{r}
ri <- psychnet(SRL_GPT, method = "relimp")
ri
```

The incoming shares per node sum to that node's R-squared; the certificate
confirms this efficiency identity holds to numerical precision:

```{r}
certificate(ri)
```

The directed edges -- each predictor's share of an outcome's variance:

```{r}
as.data.frame(ri)
```

Per-node explained variance that the incoming edges reconstruct:

```{r}
net_predict(ri)
```

## Plotting

`cograph::splot()` draws the directed network with arrows. Because each pair has
two distinct edges (`A -> B` and `B -> A`), pass `directed = TRUE` so they are
drawn separately, with `psych_styling = TRUE` (green = positive, red = negative)
and the predictability ring via `predictability = TRUE`.

```{r relimp-plot, eval = has_cograph}
cograph::splot(ri, directed = TRUE, psych_styling = TRUE, predictability = TRUE)
```
