---
title: "Visualizing networks with cograph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing networks with cograph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_cograph <- requireNamespace("cograph", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 6, dpi = 100)
library(psychnets)
```

`psychnets` belongs to the Dynalytics ecosystem: a framework for the rigorous
analytics of dynamical systems. Dynalytics treats estimation, confirmatory
testing, analysis, and software as parts of one scientific contract: every
analytical claim should be matched by evidence appropriate to the structure and
complexity of that claim. Within that contract, `psychnets` covers psychological
networks with clean-room, base-R estimators, correctness certificates, tidy
network objects, and framework verbs for centrality, predictability,
bootstrapping, stability, and group comparison.

`cograph` is the visualization layer for this ecosystem. It is built for fast,
customizable network graphics across dynamical, social, psychological,
heterogeneous, and multilayer networks. A fitted `psychnets` object can therefore
move directly into `cograph` without conversion: the same model can be rendered
with different layouts, themes, node shapes, centrality-scaled sizes, signed-edge
colors, thresholds, edge labels, group panels, and export settings. This vignette
shows that plotting workflow from the default graph to publication-oriented
figures.

```{r check, echo = FALSE, results = "asis", eval = !has_cograph}
cat("> `cograph` is not installed, so the cograph plotting chunks are skipped.")
```

# Fit Once

```{r fit, eval = has_cograph}
fit <- ebic_glasso(SRL_Claude)
fit
```

# Default cograph Plot

```{r default, eval = has_cograph}
plot(fit)
```

The explicit `splot()` call is useful when you want to set plotting parameters:

```{r splot-default, eval = has_cograph}
cograph::splot(fit)
```

# Layouts

The same network can be drawn with different layouts. Fixed seeds make stochastic
layouts reproducible.

```{r layouts, eval = has_cograph, fig.width = 9, fig.height = 4.5}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
cograph::splot(fit, layout = "circle", title = "circle")
cograph::splot(fit, layout = "spring", seed = 11, title = "spring")
par(op)
```

# Nodes and Labels

Node aesthetics can be set directly: fill, border, label size, and shape. Scaling
node size by a centrality measure makes the structural role of each node visible.

```{r nodes, eval = has_cograph}
cograph::splot(
  fit,
  layout = "circle",
  scale_nodes_by = "strength",
  node_size_range = c(4, 12),
  node_fill = c("#4C78A8", "#F58518", "#54A24B", "#B279A2", "#E45756"),
  node_border_color = "white",
  node_border_width = 2,
  label_size = 0.9,
  title = "Node size scaled by strength"
)
```

# Edges

For signed psychometric networks, edge color and width usually matter more than
decorative node styling. Thresholding and edge labels help when the complete
network is too dense.

```{r edges, eval = has_cograph}
cograph::splot(
  fit,
  layout = "circle",
  threshold = 0.05,
  edge_width_range = c(0.5, 5),
  edge_positive_color = "#2A9D8F",
  edge_negative_color = "#E76F51",
  edge_labels = TRUE,
  edge_label_size = 0.7,
  edge_label_bg = "white",
  title = "Thresholded edge weights"
)
```

The edge-weight distribution can be inspected separately:

```{r edge-weights, eval = has_cograph, fig.width = 7, fig.height = 4.5}
cograph::plot_edge_weights(fit)
```

# Bootstrap and Edge Differences

Bootstrap diagnostics are estimated by `psychnets`. The default plot shows
edge-weight confidence intervals.

```{r bootstrap, eval = has_cograph, fig.width = 8, fig.height = 5}
set.seed(1)
bs <- net_boot(SRL_Claude, method = "glasso", n_boot = 250, cores = 1)
plot(bs)
```

The same bootstrap object also contains the retained draws used for pairwise
edge-difference tests.

```{r edge-diff-box, eval = has_cograph, fig.width = 6.5, fig.height = 6.5}
plot(bs, type = "edge_diff")
```

Use the forest style when the effect sizes and confidence intervals matter more
than the significance-box display.

```{r edge-diff-forest, eval = has_cograph, fig.width = 7, fig.height = 5}
plot(difference_test(bs, type = "edge"), style = "forest")
```

# Themes

Themes apply coordinated defaults without changing the fitted model.

```{r themes, eval = has_cograph, fig.width = 9, fig.height = 4.5}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
cograph::splot(fit, theme = "minimal", title = "minimal")
cograph::splot(fit, theme = "colorblind", title = "colorblind")
par(op)
```

# Group Networks

`psychnet(..., group = )` returns a collection of fitted networks. `cograph`
plots the collection as a shared-layout grid, so node positions are comparable
across groups.

```{r group-data, include = FALSE}
grouped_srl <- rbind(
  data.frame(SRL_Claude, source = "Claude", check.names = FALSE),
  data.frame(SRL_GPT, source = "GPT", check.names = FALSE)
)
```

```{r groups, eval = has_cograph, fig.width = 8, fig.height = 4.5}
group_fit <- psychnet(grouped_srl, group = "source", method = "glasso")
cograph::splot(group_fit, layout = "circle", psych_styling = TRUE)
```

# Export

Use `filename`, `width`, `height`, and `res` when a figure is ready to save:

```{r export, eval = FALSE}
cograph::splot(fit, layout = "circle", filename = "network.png",
               width = 8, height = 8, res = 300)
```

# Useful `splot()` Options

| Task | Arguments |
|---|---|
| Layout | `layout`, `seed`, `layout_scale`, `layout_margin` |
| Nodes | `node_size`, `scale_nodes_by`, `node_size_range`, `node_fill`, `node_shape` |
| Labels | `labels`, `label_size`, `label_color`, `label_position` |
| Edges | `edge_width_range`, `edge_positive_color`, `edge_negative_color`, `edge_alpha` |
| Filtering | `threshold`, `minimum`, `maximum` |
| Edge labels | `edge_labels`, `edge_label_size`, `edge_label_bg`, `edge_label_style` |
| Themes | `theme`, `background`, `title` |
| Export | `filename`, `width`, `height`, `res` |
