---
title: "Visualizing bootstrap, centrality and difference results (base R)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing bootstrap, centrality and difference results (base R)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5, dpi = 100)
library(psychnets)
```

`psychnets` ships **native base-R plot methods** for every diagnostic verb, so
you can inspect a model's stability and structure with no additional packages
(`graphics` and `grDevices` ship with R). This vignette walks through them on the
bundled `SRL_Claude` self-regulated-learning data (300 respondents, five MSLQ
subscales). A companion vignette,
*"Visualizing networks with cograph"*, shows the richer cograph renderings of the
same results.

Every plot is an S3 `plot()` method on a result object — call `plot()` on what a
verb returns; there is no bracket-subsetting to assemble.

# Estimate a network

```{r fit}
X   <- as.matrix(SRL_Claude)
fit <- ebic_glasso(X)
fit
```

# Centralities

`net_centralities()` returns a tidy table; `plot()` draws it.

```{r centralities}
ct <- net_centralities(fit, measures = c("strength", "expected_influence", "betweenness"))
ct
```

`type = "bar"` (the default) gives one sorted lollipop panel per measure:

```{r cent-bar, fig.width = 9, fig.height = 4.5}
plot(ct)
```

`type = "line"` gives the qgraph-style faceted centrality plot — one panel per
measure, nodes sharing a single vertical order, each panel on its own (raw)
axis:

```{r cent-line, fig.width = 8, fig.height = 4.5}
plot(ct, type = "line")
```

Use `scale = "z"` or `scale = "relative"` if you prefer a standardised or a
0--1 axis instead of raw values.

# Bootstrap accuracy

`net_boot()` resamples and re-estimates; `plot()` defaults to the edge-weight
confidence intervals, sorted by the observed weight (edges whose interval
excludes zero are emphasised).

```{r boot}
set.seed(1)
bs <- net_boot(X, method = "glasso", n_boot = 250, cores = 1)
plot(bs)
```

Bootstrapped centrality intervals, one panel per measure:

```{r boot-cent, fig.width = 8, fig.height = 5}
plot(bs, type = "centrality")
```

The bootstrapped difference "significance box" matrices — which edges, or which
node strengths, differ from one another:

```{r boot-diff, fig.width = 6.5, fig.height = 6.5}
plot(bs, type = "edge_diff")
plot(bs, type = "centrality_diff", measure = "strength")
```

# Difference test

`difference_test()` is the tidy table of pairwise differences. Its `plot()` has
two styles: the significance-box matrix (`"box"`, the default) for *which* pairs
differ, and a forest plot (`"forest"`) for *how big* each difference is with its
confidence interval.

```{r difftest, fig.width = 6.5, fig.height = 6.5}
dt <- difference_test(bs, type = "strength")
plot(dt)                       # box matrix
```

```{r difftest-forest, fig.width = 7, fig.height = 5}
plot(dt, style = "forest")     # forest plot
```

# Case-dropping stability

`net_stability()` drops increasing fractions of cases and correlates the
subset centralities with the full-sample ones; `plot()` shows the curves, a
±1 SD band, the acceptance threshold, and the CS-coefficient per measure.

```{r stability, fig.width = 7, fig.height = 5}
st <- net_stability(X, method = "glasso",
                    drop_prop = seq(0.1, 0.8, 0.1), iter = 25)
plot(st)
```

# Network comparison test

`net_compare()` permutation-tests two networks. Plot the global-strength (M)
and structure (S) permutation nulls, or the per-edge differences.

```{r nct}
cmp <- net_compare(X, as.matrix(SRL_GPT), iter = 250)
plot(cmp)                      # global strength (M)
```

```{r nct-edges, fig.width = 7, fig.height = 5}
plot(cmp, type = "edges")      # per-edge differences
```

# Summary

| Result object (verb) | `plot()` options |
|---|---|
| `net_centralities()` | `type = "bar"` / `"line"`; `scale = "raw"`/`"z"`/`"relative"` |
| `net_boot()` | `type = "edges"`/`"centrality"`/`"edge_diff"`/`"centrality_diff"`/`"predictability"` |
| `difference_test()` | `style = "box"`/`"forest"` |
| `net_stability()` | — |
| `net_compare()` | `type = "strength"`/`"structure"`/`"edges"` |

All of the above use only base R. For the estimated network itself, `plot(fit)`
delegates to `cograph::splot()` — see the companion cograph vignette.
