---
title: "Visualizing bootstrap, centrality and difference results (base R)"
author: "[Mohammed Saqr](https://saqr.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing bootstrap, centrality and difference results (base R)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5, dpi = 100)
library(psychnets)
```

## Reading a network diagnostic as a picture

A psychometric network is estimated from a finite sample, so every edge weight
and every centrality value carries sampling uncertainty. The diagnostic verbs in
`psychnets` quantify that uncertainty by resampling, case-dropping, and
permutation. Each verb returns a tidy result object, and each result object
carries a base-`graphics` `plot()` method that turns the numbers into a figure.
The plots are drawn with the `graphics` and `grDevices` engines that ship with
R, so reading a diagnostic needs no additional package.

Every figure in this vignette is the `plot()` method for one result class. The
pattern is uniform: fit a network, call a diagnostic verb, then call `plot()` on
what the verb returns. The sections below cover the centrality plot, the
bootstrap plots and their difference-box matrices, the pairwise difference test,
the case-dropping stability curves, and the network comparison test.

## The data

The bundled `SRL_GPT` data hold responses from 300 learners on five
self-regulated-learning constructs (CSU, IV, SE, SR, TA), each a continuous
subscale score.

```{r data}
head(SRL_GPT)
```

## Estimate a network

`ebic_glasso()` estimates a Gaussian graphical model whose edges are
partial correlations selected by the extended Bayesian information criterion. The
fitted object is the input to every diagnostic below.

```{r fit}
fit <- ebic_glasso(SRL_GPT)
fit
```

## Centralities

`net_centralities()` computes node centrality measures from the fitted
network and returns a tidy table with one row per node and one column per
measure. Strength is the sum of absolute edge weights at a node, expected
influence is the sum of signed edge weights, and betweenness counts the weighted
shortest paths that pass through a node.

```{r centralities}
ct <- net_centralities(fit, measures = c("strength", "expected_influence", "betweenness"))
ct
```

`plot()` for a `psychnet_centrality` table draws one panel per measure.
Under the default `type = "bar"`, each panel is a horizontal lollipop chart whose
nodes are sorted by their value, so the most central node sits at the top of the
panel and the length of each segment reads directly as the centrality value.

```{r cent-bar, fig.width = 9, fig.height = 4.5}
plot(ct)
```

Under `type = "line"`, `plot()` draws the faceted centrality profile.
The nodes share a single vertical order across all panels, set by their mean
standardized centrality, and each panel keeps its own horizontal axis. A node
that sits high in every panel is central on every measure; a node whose position
shifts between panels is central on one measure and peripheral on another.

```{r cent-line, fig.width = 8, fig.height = 4.5}
plot(ct, type = "line")
```

The `scale` argument controls the per-panel transform in the line view. The
default `"raw"` keeps each panel on its own axis of centrality values, `"z"`
centres each measure at its mean and scales it to unit standard deviation, and
`"relative"` maps each measure to the interval from zero to one.

```{r cent-line-z, fig.width = 8, fig.height = 4.5}
plot(ct, type = "line", scale = "z")
```

## Bootstrap accuracy

`net_boot()` resamples the rows of the data with replacement, refits the
network on each resample, and records the distribution of every edge weight and
every centrality value across resamples. The percentile confidence interval of
each quantity is the reported measure of its sampling accuracy.

```{r boot}
set.seed(1)
bs <- net_boot(SRL_GPT, method = "glasso", n_boot = 250, cores = 1)
```

`plot()` for a `psychnet_bootstrap` object defaults to `type = "edges"`,
the edge-accuracy plot. Each row is an edge, sorted by its observed weight; the
point is the observed weight and the segment is its bootstrap confidence
interval. An edge whose interval excludes zero is drawn in the emphasis colour,
which reports that the edge is present across resamples; an edge whose interval
crosses zero is muted.

```{r boot-edges}
plot(bs)
```

Under `type = "centrality"`, `plot()` draws the bootstrapped centrality
intervals, one sorted panel per measure. A node whose interval is wide relative
to the spread across nodes is estimated with low precision, so its rank among the
nodes is uncertain.

```{r boot-cent, fig.width = 8, fig.height = 5}
plot(bs, type = "centrality")
```

Under `type = "edge_diff"` and `type = "centrality_diff"`, `plot()`
draws the difference "significance box" matrix. The items sit on both axes,
ordered by their observed value from high to low. The diagonal carries the
observed value, shaded by its magnitude. An off-diagonal cell is filled when that
pair of items differs across the bootstrap, coloured red when the row item is
larger than the column item and blue when it is smaller, and it is left faint
when the pair does not differ. The stars report the difference p-value.

```{r boot-diff, fig.width = 6.5, fig.height = 6.5}
plot(bs, type = "edge_diff")
```

```{r boot-cent-diff, fig.width = 6.5, fig.height = 6.5}
plot(bs, type = "centrality_diff", measure = "strength")
```

## Difference test

`difference_test()` reads the retained bootstrap draws and returns a
tidy table of pairwise differences, with columns `item1`, `item2`, the two
observed values, the observed difference `obs_diff`, its interval `lower` and
`upper`, the `p_value`, and a `significant` flag set when the interval excludes
zero.

```{r difftest-tab}
dt <- difference_test(bs, type = "strength")
head(dt)
```

`plot()` for a `psychnet_difference` table has two styles. The default
`style = "box"` draws the same significance-box matrix as above, which reads
*which* pairs differ.

```{r difftest-box, fig.width = 6.5, fig.height = 6.5}
plot(dt)
```

Under `style = "forest"`, `plot()` draws a forest plot with one row per
pair. The point is the observed difference, the segment is its confidence
interval, the dashed line marks zero, and a pair whose interval excludes zero is
emphasised. This view reads *how large* each difference is together with its
uncertainty.

```{r difftest-forest, fig.width = 7, fig.height = 5}
plot(dt, style = "forest")
```

## Case-dropping stability

`net_stability()` drops an increasing fraction of the cases, refits the
network on each subset, and correlates the subset centralities with the
full-sample centralities. The case-dropping (CS) coefficient is the largest drop
proportion at which the correlation stays above the acceptance threshold with the
stated certainty.

```{r stability-fit}
st <- net_stability(SRL_GPT, method = "glasso",
                    drop_prop = seq(0.1, 0.8, 0.1), iter = 25)
```

`plot()` for a `psychnet_stability` object draws the stability curves.
The horizontal axis is the proportion of cases dropped and the vertical axis is
the mean correlation with the full sample. Each measure is a line with a band of
plus or minus one standard deviation, the dashed line is the acceptance
threshold, and the legend reports the CS-coefficient for each measure. A curve
that stays above the threshold as more cases are dropped indicates a stable
centrality ordering.

```{r stability, fig.width = 7, fig.height = 5}
plot(st)
```

## Network comparison test

`net_compare()` compares two networks by permutation. It repeatedly
reassigns the pooled cases to two groups at random, recomputes the test statistic
on each permutation, and builds the distribution of that statistic expected when
the two networks do not differ. The global strength statistic (M) is the
difference in summed absolute edge weights, and the structure statistic (S) is
the maximum absolute edge difference.

```{r nct-fit}
cmp <- net_compare(SRL_GPT, SRL_Claude, iter = 250)
```

`plot()` for a `psychnet_nct` object defaults to `type = "strength"`,
the permutation null for M. The histogram is the null distribution, the red line
marks the observed value, and the title reports the permutation p-value. An
observed value in the tail of the null provides evidence that the two networks
differ in global strength.

```{r nct, fig.width = 7, fig.height = 5}
plot(cmp)
```

Under `type = "structure"`, `plot()` draws the same null for the maximum
edge-difference statistic S.

```{r nct-structure, fig.width = 7, fig.height = 5}
plot(cmp, type = "structure")
```

Under `type = "edges"`, `plot()` draws the observed absolute difference
of each edge as a horizontal bar, coloured by whether that edge differs
significantly at the level set by `alpha`. The title reports how many edges reach
significance.

```{r nct-edges, fig.width = 7, fig.height = 5}
plot(cmp, type = "edges")
```

## The plot methods at a glance

| Result object (verb) | `plot()` options |
|---|---|
| `net_centralities()` | `type = "bar"` / `"line"`; `scale = "raw"` / `"z"` / `"relative"` |
| `net_boot()` | `type = "edges"` / `"centrality"` / `"edge_diff"` / `"centrality_diff"` / `"predictability"` |
| `difference_test()` | `style = "box"` / `"forest"` |
| `net_stability()` | one figure |
| `net_compare()` | `type = "strength"` / `"structure"` / `"edges"` |

Each figure above is drawn with base R alone. For the estimated network itself,
`plot(fit)` delegates to `cograph::splot()`, covered in the companion vignette
*Visualizing networks with cograph*.
