---
title: "Edge-weight stability and split-half reliability"
author: "[Mohammed Saqr](https://saqr.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Edge-weight stability and split-half reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5.5, dpi = 100)
library(psychnets)
set.seed(1)
```

## What network accuracy and stability are

A psychometric network is an estimate. Every edge weight and every centrality
value is computed from a finite sample, so each carries sampling variability. A
network drawn from one sample would differ from a network drawn from another
sample of the same population. Reliability analysis is the set of resampling
diagnostics that quantify this variability and report how much of the estimated
structure is reproducible.

Four questions organise the diagnostics in this vignette. How wide is the
sampling interval around each edge weight, and does that interval exclude zero?
Do the differences between edges or between centralities survive resampling? Do
the centrality rankings hold up when cases are dropped? Does the whole edge
structure reproduce across independent halves of the sample? The verbs
`net_boot()`, `difference_test()`, `net_stability()`, `casedrop_reliability()`,
and `network_reliability()` answer these in turn. Every verb is
estimator-agnostic: each routes its refits through `psychnet(method = )`, so the
same diagnostic applies to any estimator in the package, and each is written in
base R with no compiled dependency.

## The data

The bundled `SRL_Claude` data set holds 300 respondents scored on five
self-regulated-learning subscales: cognitive strategy use (CSU), intrinsic value
(IV), self-efficacy (SE), self-regulation (SR), and test anxiety (TA).

```{r data}
head(SRL_Claude)
```

## Bootstrapped edge and centrality accuracy

`net_boot()` resamples the observations with replacement, re-estimates
the network on each resample, and summarises the sampling distribution of every
edge weight and every node centrality. For each edge it reports the resample
mean, the percentile confidence interval, the proportion of resamples in which
the edge is non-zero, and a `significant` flag that is `TRUE` when the interval
excludes zero. The returned object has class `psychnet_bootstrap`; its `$edges`
data frame carries the columns `from`, `to`, `observed`, `mean`, `lower`,
`upper`, `prop_nonzero`, and `significant`, and its `$centrality` data frame
carries the observed centrality of each node with a lower and upper bound per
measure.

```{r boot}
b <- net_boot(SRL_Claude, method = "glasso", n_boot = 200, cores = 1)
b
```

The header reports 200 resamples, a 95% interval, and 8 of the 10 edges flagged
significant. The edge table gives the accuracy interval behind each estimate.

```{r boot-edges}
b$edges
```

A bootstrap confidence interval is the range of edge weights consistent with the
data at the stated level. An edge whose interval excludes zero is a non-zero
edge. The CSU–SR edge has an observed weight of 0.73 and an interval of about
0.65 to 0.77, well clear of zero, so it is one of the reliably non-zero edges.
The CSU–IV edge has an observed weight near 0.002 and an interval from 0 to
about 0.12, which includes zero, so its sign and presence are undetermined at
this sample size; the IV–TA edge behaves the same way. The `prop_nonzero`
column reads these two edges as selected in only a part of the resamples (about
0.55 and 0.76), while the eight significant edges are selected in every
resample.

The default plot draws each edge weight with its bootstrap interval, ordered by
magnitude, with the significant edges marked.

```{r boot-plot, fig.width = 7, fig.height = 6}
plot(b)
```

The centrality panel does the same for the node measures.

```{r boot-cent-plot, fig.width = 7, fig.height = 5}
plot(b, type = "centrality")
```

## Within-network difference tests

`difference_test()` tests, inside a single network, whether two edge
weights or two node centralities differ. For each pair it forms the
per-resample difference from the stored bootstrap draws, takes the percentile
interval of that difference, and flags the pair `significant` when the interval
excludes zero; it also reports the two-sided bootstrap p-value. The verb reads
the draws already stored on a `net_boot()` object, so no refitting happens. It
returns a tidy data frame with columns `item1`, `item2`, `value1`, `value2`,
`obs_diff`, `lower`, `upper`, `p_value`, and `significant`. This is the
within-network counterpart to the accuracy intervals of `net_boot()`: it asks
which nodes are ordered with confidence.

```{r difference}
difference_test(b, type = "strength")
```

A two-sided bootstrap p-value is the probability, under resampling, of a
difference at least as large as the observed one when the two quantities are
equal. Self-regulation (SR) has the highest observed strength at 1.95, and its
difference from every other node has an interval excluding zero and a p-value of
0.00, so SR ranks above the rest with confidence. The SE–TA pair is the
exception: its observed strength difference is about -0.12, its interval runs
from about -0.33 to 0.08 and includes zero, and its p-value is 0.26, so the
strengths of SE and TA are not distinguishable here. A ranking that separates
those two nodes would not be supported by the data.

## Centrality stability under case-dropping

`net_stability()` re-estimates the network on random case-dropped
subsets and correlates each subset's centralities, by Spearman rank, with the
full-sample centralities. The correlation-stability coefficient, or
CS-coefficient, is the largest proportion of cases that can be dropped while the
rank correlation with the full-sample centrality stays at or above a threshold
(0.7 by default) with a stated probability (0.95 by default). The returned
`psychnet_stability` object holds `$cs`, the CS-coefficient per measure, and a
tidy `$table` with columns `measure`, `drop_prop`, `mean_cor`, `sd_cor`, and
`prop_above`.

```{r stability}
st <- net_stability(SRL_Claude, method = "glasso", iter = 100)
st
```

The CS-coefficient for strength is 0.60 and for expected influence is 0.90. The
interpretation of a CS-coefficient of 0.60 is that up to 60% of the cases can be
dropped and the strength ranking still correlates at least 0.7 with the
full-sample ranking in 95% of subsets. A common rule of thumb sets 0.25 as the
minimum and 0.50 as the preferred value for a centrality ordering that can be
read with confidence; both measures here sit above 0.50, and expected influence
is close to the maximum. The plot draws the mean correlation against the drop
proportion with a plus-or-minus one standard deviation band and the acceptance
threshold.

```{r stability-plot, fig.width = 7, fig.height = 5}
plot(st)
```

## Edge-weight case-dropping stability

`casedrop_reliability()` provides the edge-vector complement of
`net_stability()`. For each drop proportion it re-estimates the network on
random case-dropped subsets and compares the subset edge-weight vector with the
full-sample one, reporting the correlation between the two vectors together with
the mean, median, and maximum absolute edge deviation. The edge-weight
CS-coefficient is the largest drop proportion at which the edge-vector
correlation stays at or above the threshold with the stated probability. The
result is a tidy data frame of class `psychnet_casedrop`, one row per metric per
drop proportion, with columns `metric`, `drop_prop`, `mean`, and `sd`; the
CS-coefficient prints on the header line.

```{r casedrop}
cd <- casedrop_reliability(SRL_Claude, method = "glasso", iter = 100)
cd
```

The edge-weight CS-coefficient is 0.90. The `correlation` rows read the vector
correlation directly: at a 10% drop the subset edge vector correlates 0.98 with
the full-sample vector, and even at a 90% drop the correlation holds near 0.92,
so the edge structure as a whole is highly reproducible under case-dropping. The
absolute-deviation rows report the size of the edge shifts on the weight scale:
the mean absolute deviation grows from about 0.012 at a 10% drop to about 0.11
at a 90% drop, and the maximum single-edge deviation from about 0.03 to about
0.26. The four metrics plot against the drop proportion, each with a
plus-or-minus one standard deviation band; the correlation panel carries the
threshold and the CS-coefficient.

```{r casedrop-plot, fig.width = 8, fig.height = 6.5}
plot(cd)
```

## Split-half reliability

`network_reliability()` repeatedly splits the sample into two halves,
estimates a network on each half, and compares the two edge-weight vectors. It
reports, across the splits, the edge-weight correlation between halves together
with the mean, median, and maximum absolute edge deviation, a psychometric
reliability view of the estimated structure. The result is a tidy data frame of
class `psychnet_reliability`, one row per metric, with columns `metric`, `mean`,
`sd`, `lower`, and `upper`, where `lower` and `upper` are the 2.5% and 97.5%
quantiles of the metric across the split-half iterations.

```{r reliability}
rel <- network_reliability(SRL_Claude, method = "glasso", iter = 100)
rel
```

The correlation between the two independent halves averages 0.982 with a 95%
range from about 0.953 to 0.996 across the 100 splits. A between-halves edge
correlation near 1 indicates that two disjoint samples of 150 respondents each
recover almost the same edge structure. The mean absolute edge deviation
averages 0.064 and the maximum single-edge deviation averages 0.142, so the
typical edge shifts by well under a tenth of a unit between halves while the
largest shift reaches about a seventh. The plot shows the distribution of each
metric across the splits with the observed mean marked.

```{r reliability-plot, fig.width = 8, fig.height = 6.5}
plot(rel)
```

## Any estimator, any grouping

Every verb in this vignette routes its refits through `psychnet(method = )`, so
each applies to any estimator with no per-method code. The partial-correlation
estimator gives its own case-dropping profile, for example.

```{r pcor}
casedrop_reliability(SRL_Claude, method = "pcor", iter = 100)
```

Each verb also accepts a `psychnet_group` object, built with
`psychnet(..., group = )`, and returns one result per group level.

## Summary

| Verb | Question it answers | Returns |
|---|---|---|
| `net_boot()` | How wide is the accuracy interval around each edge and centrality? | `psychnet_bootstrap` with tidy `$edges` and `$centrality` |
| `difference_test()` | Do two edges or two centralities differ within the network? | tidy df (`item1`, `item2`, interval, `p_value`, `significant`) |
| `net_stability()` | How robust is the centrality ranking to dropping cases? | `psychnet_stability` with `$cs` and `$table` |
| `casedrop_reliability()` | How robust is the edge structure to dropping cases? | tidy df (metric by drop proportion) with the edge-weight CS |
| `network_reliability()` | How reproducible is the edge structure across split-halves? | tidy df, one row per between-halves metric |

Together these verbs cover edge accuracy, within-network ordering, centrality
stability, edge-weight stability, and split-half reliability in base R.
